New cardinality estimation algorithms for HyperLogLog sketches

### 2016-06-19 paper/paper.tex
improvements

% SYMBOLS
\newcommand*{\symDefine}[2]{\newcommand{{#1}}{{#2}}}
\symDefine{\symPrecision}{p}
\symDefine{\symRegRange}{q}
\symDefine{\symRegVal}{k}
\symDefine{\symRegValVariate}{K}
\symDefine{\symNumReg}{m}
\symDefine{\symDataItem}{D}
\symDefine{\symBitRepA}{a}
\symDefine{\symBitRepB}{b}
\symDefine{\symIndexI}{i}
\symDefine{\symIndexK}{k}
\symDefine{\symIndexL}{l}
\symDefine{\symIndexJ}{j}
\symDefine{\symCount}{c}
\symDefine{\symAlpha}{\alpha}
\symDefine{\symBeta}{\beta}
\symDefine{\symA}{a}
\symDefine{\symB}{b}
\symDefine{\symE}{e}
\symDefine{\symX}{x}
\symDefine{\symXEstimate}{\hat{\symX}}
\symDefine{\symXNormalized}{\symX'}
\symDefine{\symN}{n}
\symDefine{\symS}{s}
\symDefine{\symY}{y}
\symDefine{\symZ}{z}
\symDefine{\symInteger}{z}
\symDefine{\symIntegerVariate}{Z}
\symDefine{\symCardinality}{n}
\symDefine{\symCardinalityEstimate}{\hat{\symCardinality}}
\symDefine{\symCardinalityOriginalEstimate}{\symCardinalityEstimate_\text{raw}}
\symDefine{\symError}{\varepsilon}
\symDefine{\symStopDelta}{\delta}
\symDefine{\symStopEpsilon}{\varepsilon}
\symDefine{\symBigO}{\mathcal{O}}
\symDefine{\symExpectation}{\mathbb{E}}
\symDefine{\symProbability}{P}
\symDefine{\symProbabilityMass}{\rho}
\symDefine{\symRegProbability}{\gamma}
\symDefine{\symNu}{\nu}
\symDefine{\symKappa}{\kappa}
\symDefine{\symLikelihood}{\mathcal{L}}
\symDefine{\symPoissonRate}{\lambda}
\symDefine{\symPoissonRateEstimate}{\hat{\symPoissonRate}}
\symDefine{\symFunc}{f}
\symDefine{\symFuncPrime}{g}
\symDefine{\symHelper}{h}
\symDefine{\symHelperApprox}{\tilde{\symHelper}}
\symDefine{\symSetA}{A}
\symDefine{\symSetB}{B}
\symDefine{\symSetS}{S}
\symDefine{\symSetX}{X}
\symDefine{\symSetASuffix}{a}
\symDefine{\symSetBSuffix}{b}
\symDefine{\symSetXSuffix}{x}
\symDefine{\symCountMatrix}{\mathbf{c}}
\symDefine{\symPowerSeriesFunc}{\xi}
\symDefine{\symSmallCorrectionFunc}{\sigma}
\symDefine{\symLargeCorrectionFunc}{\tau}
\symDefine{\symEpsPowerSeriesFunc}{\varepsilon_\symPowerSeriesFunc}

@@

In the history of the HyperLogLog algorithm different averaging techniques have been proposed. First, there was the LogLog algorithm using the geometric mean and the SuperLogLog algorithm that enhanced the estimate by truncating the largest register values before applying the geometric mean \cite{Durand2003}. Finally, the harmonic mean was found to give even better estimates as it is inherently less sensitive to outliers. The result is the so-called raw estimator and given by
\begin{equation}
\label{equ:intermediate_estimate}
\symCardinalityOriginalEstimate
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexJ=1}^{\symNumReg}2^{-\symRegVal_\symIndexJ}}
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCount_\symRegVal 2^{-\symRegVal}}
\end{equation}
where $\symAlpha_\symNumReg$  is a correction factor that is fixed for a given number of registers $\symNumReg$ and which was derived to be \cite{Flajolet2007}
\begin{equation}
\symAlpha_\symNumReg := \left(
\symNumReg
\int_0^\infty
\left(
\log_2\left(
\frac{2+u}{1+u}
\right)
\right)^\symNumReg
du
\right)^{-1}.
\end{equation}
Numerical approximations of $\symAlpha_\symNumReg$ for various values of $\symNumReg$ have been listed in \cite{Flajolet2007}. These approximations are used in many HyperLogLog implementations. However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions $\symPrecision$. For HyperLogLog sketches that are used in practice with 256 or more registers ($\symPrecision\geq 8$), it is completely sufficient to use 
\begin{equation}
\label{equ:alpha_inf}
\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg = \frac{1}{2\log 2} \approx 0.7213475,
\end{equation}
as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:intermediate_estimate}, because the additional bias is negligible compared to the estimation error.

@@

As mentioned in \cref{sec:data_element_insertion}, a HyperLogLog sketch with parameters $(\symPrecision, \symRegRange)$ can be turned into a $(\symPrecision, 0)$-HyperLogLog sketch. Since $\symRegRange=0$ corresponds to linear counting and the transformed HyperLogLog sketch corresponds to a bitset with $\symCount_0$ zeros, the linear counting cardinality estimator \cite{Whang1990} can be applied here
\begin{equation}
\label{equ:small_range_estimate}
\symCardinalityEstimate_\text{small} = \symNumReg \log(\symNumReg/\symCount_0).
\end{equation}
The corresponding relative estimation error as depicted in \cref{fig:small_range_estimate} shows that this estimator is convenient for small cardinalities. It was proposed to use this small-range estimator as long as $\symCardinalityOriginalEstimate \leq \frac{5}{2}\symNumReg$ where the factor $\frac{5}{2}$ was empirically determined \cite{Flajolet2007}.

For large cardinalities in the order of $2^{\symPrecision+\symRegRange}$, for which a lot of registers are already in a saturated state meaning that they have reached the maximum possible value $\symRegRange + 1$, the raw estimator underestimates the cardinalities. For the 32-bit hash value case $(\symPrecision+\symRegRange=32)$ which was considered in \cite{Flajolet2007}, following correction formula was proposed to take these saturated registers into account
\begin{equation}
\label{equ:large_range_estimate}
\symCardinalityEstimate_\text{large}
=
-2^{32}\log\!\left(1-\symCardinalityOriginalEstimate/2^{32}\right).
\end{equation}
The original estimation algorithm as presented in \cite{Flajolet2007} including small and large range corrections is summarized by \cref{alg:estimate_original}. 
\begin{algorithm}
\caption{Original cardinality estimation algorithm for HyperLogLog sketches that use 32-bit hash values ($\symPrecision+\symRegRange = 32$) for insertion of data items \cite{Flajolet2007}.}
\label{alg:estimate_original}
\begin{algorithmic}
\Function {EstimateCardinality}{$\vec{\symRegVal}$}
\State $\symNumReg \gets \dim \vec{\symRegVal}$
\State $\symCardinalityOriginalEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexJ=1}^\symNumReg 2^{-\symRegVal_\symIndexJ}\right)^{-1}$
\Comment raw estimate \eqref{equ:intermediate_estimate}
\If{$\symCardinalityOriginalEstimate\leq \frac{5}{2}\symNumReg$}
\State $\symCount_0 = \left|\{\symIndexJ |\symRegVal_\symIndexJ=0\}\right|$
\If{$\symCount_0\neq0$}
\State \Return $\symNumReg \log(\symNumReg / \symCount_0)$ \Comment small range correction \eqref{equ:small_range_estimate}
\Else
\State \Return $\symCardinalityOriginalEstimate$
\EndIf
\ElsIf{$\symCardinalityOriginalEstimate\leq \frac{1}{30}2^{32}$}
\State \Return $\symCardinalityOriginalEstimate$
\Else
\State\Return $-2^{32}\log(1-\symCardinalityOriginalEstimate/2^{32})$\Comment large range correction \eqref{equ:large_range_estimate}
\EndIf
\EndFunction
\end{algorithmic}
\end{algorithm}
The relative estimation error for the original method is shown in \cref{fig:original_estimate}. Unfortunatley, as can be clearly seen, the ranges where the estimation error is small for $\symCardinalityOriginalEstimate$ and $\symCardinalityEstimate_\text{small}$ do not overlap. Therefore, the estimation error is much larger near the transition region. To reduce the estimation error for cardinalities close to this region, it was proposed to correct $\symCardinalityOriginalEstimate$ for bias. Empirically collected bias correction data is either stored as set of interpolation points \cite{Heule2013}, as lookup table \cite{Rhodes2015}, or as best-fitting polynomial \cite{Sanfilippo2014}. However, all these empirical approaches treat the symptom and not the cause.

@@

The large range correction is not satisfying either as it does not reduce the estimation error. Quite the contrary, it even increases the estimation error. However, instead of underestimating the cardinalities, they are now overestimated. Another indication for the incorrectness of the proposed large range correction \eqref{equ:large_range_estimate} is the fact that it is not even defined for all possible states. For instance, consider a $(\symPrecision,\symRegRange)$-HyperLogLog sketch with $\symPrecision+\symRegRange=32$ for which all registers are equal to the maximum possible value $\symRegRange+1$. The raw estimate would be $\symCardinalityEstimate_\text{raw} = \symAlpha_\symNumReg 2^{33}$, which is greater than $2^{32}$ and outside of the domain of the large range correction.

A simple approach to avoid the large range correction at all is the use of larger hash values $(\symPrecision+\symRegRange>32)$ which shift the biased region of $\symCardinalityOriginalEstimate$ to higher cardinalities. However, in case $\symRegRange \geq 30$ 6-bit registers as used in \cite{Heule2013} are required to allow representing all values in the range $[0, \symRegRange+1]$. If 64-bit hash values are used as proposed in \cite{Heule2013} $(\symPrecision+\symRegRange=64)$, a correction for large cardinalities can be circumvented. Cardinalities in the order of $2^{64}$ for which saturated registers become a problem never occur in practice.

@@

Consider following integer distribution
\begin{equation}
\label{equ:assumed_register_val_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate) = e^{-\frac{\symPoissonRate}{\symNumReg 2^{ \symRegVal}}}.
\end{equation}
This distribution has infinite support and differs from the register value distribution under the Poisson model \eqref{equ:register_value_distribution}, whose support is limited to the range $[0, \symRegRange+1]$. The expectation of $2^{-\symRegVal}$ is given by
\begin{multline}
\label{equ:expectation_power_two}
\symExpectation(2^{-\symRegVal})
=
\\
\sum_{\symRegVal = -\infty}^\infty
2^{-\symRegVal}
e^{-\frac{\symPoissonRate}{\symNumReg} 2^{-\symRegVal}}
\left(1-e^{-\frac{\symPoissonRate}{\symNumReg} 2^{-\symRegVal}}\right)
=
\frac{1}{2}
\sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal}
e^{-\frac{\symPoissonRate}{\symNumReg} 2^{\symRegVal}}
=
\frac{\symAlpha_\infty\,\symNumReg\,\symPowerSeriesFunc\!\left(\log_2\!\left( \frac{\symPoissonRate}{\symNumReg}\right)\right)}{\symPoissonRate},
\end{multline}
where the function 
\begin{equation}
\label{equ:power_series_function}
\symPowerSeriesFunc(\symX):= \log(2) \sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal+\symX}
e^{-2^{\symRegVal+\symX}}
\end{equation}
 is a smooth periodic function with period 1. Numerical evaluations indicate that this function can be bounded by $1 - \symEpsPowerSeriesFunc 
\leq \symPowerSeriesFunc(\symX) \leq 1 + \symEpsPowerSeriesFunc$ with $\symEpsPowerSeriesFunc:=9.885\cdot 10^{-6}$ (see \cref{fig:power_series_func}).

@@

Let $\symRegVal_1,\ldots,\symRegVal_\symNumReg$ be a sample taken from \eqref{equ:assumed_register_val_distribution}. For large sample sizes $\symNumReg\rightarrow \infty$ we have asymptotically
\begin{equation}
\symExpectation\left(
\frac{1}{2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}}
\right)
\underset{\symNumReg\rightarrow \infty}{=}
\frac{1}{\symExpectation
\left(2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}\right)}
=
\frac{1}{\symNumReg \symExpectation\left(2^{-\symRegVal}\right)}.
\end{equation}
Together with \eqref{equ:expectation_power_two} we obtain
\begin{equation}
\symPoissonRate
=
\symExpectation\left(
\frac{\symAlpha_\infty\,\symNumReg^2\,\symPowerSeriesFunc\!\left(\log_2\!\left( \symPoissonRate/\symNumReg\right)\right)}{2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}}
\right)
\quad
\text{for $\symNumReg\rightarrow \infty$}.
\end{equation}
Therefore, 
\begin{equation}
\label{equ:rawestimator_2}
\symPoissonRateEstimate 
= 
\frac{\symAlpha_\infty\,\symNumReg^2}{2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}}
\end{equation}
is an asymptotically almost unbiased estimator for the Poisson parameter. Its asymptotic relative bias is bounded by $\symEpsPowerSeriesFunc$. This estimator corresponds to the raw estimator \eqref{equ:intermediate_estimate}, if the Poisson parameter estimate is used as cardinality estimate (see \cref{sec:poisson_approximation}).

@@

\subsection{Limitations of the raw estimator}
As we have already seen in \cref{fig:raw_estimate}, the raw estimator does not work very well for small or large cardinalities. The reason is that the distribution of register values \eqref{equ:register_value_distribution} is different from \eqref{equ:assumed_register_val_distribution} for which the raw estimator was derived. Since a random variable $\symRegValVariate'$ that obeys \eqref{equ:assumed_register_val_distribution} can be transformed into random variable $\symRegValVariate$ that follows \eqref{equ:register_value_distribution} using the transformation $\symRegValVariate = \min\left(\max\left(\symRegValVariate', 0\right), \symRegRange+1\right)$, the register values of a HyperLogLog sketch can be seen as the result after applying this transformation to a sample that is distributed according to \eqref{equ:assumed_register_val_distribution}. The corresponding multiplicities have  following relationships
\begin{equation}
\label{equ:multiplicity_transformation}
\begin{aligned}
\symCount_0 &= \textstyle\sum_{\symRegVal = -\infty}^{0} \symCount'_\symRegVal, \\
\symCount_\symRegVal &=  \symCount'_\symRegVal, \quad 1\leq\symRegVal\leq\symRegRange,\\
\symCount_{\symRegRange+1} &= \textstyle\sum_{\symRegVal = \symRegRange + 1}^{\infty} \symCount'_\symRegVal.
\end{aligned}
\end{equation}
Obviously, as long as $\symCount_0\approx\symCount'_0$ and  $\symCount'_{\symRegRange+1}\approx\symCount_{\symRegRange+1}$ the set of register values is similar to the originating sample. In case $\symCount_0=0$ and $\symCount_{\symRegRange+1}=0$, which implies $\symCount_\symRegVal=0$ for $\symRegVal\leq0$ and $\symRegVal\geq\symRegRange+1$, we can even be sure that the register values are identical to the originating sample. However, for small and large cardinalities $\symCount_0$ and $\symCount_{\symRegRange+1}$ approach $\symNumReg$, respectively, and the raw estimator needs to be corrected.

>>>>

% SYMBOLS
\newcommand*{\symDefine}[2]{\newcommand{{#1}}{{#2}}}
\symDefine{\symPrecision}{p}
\symDefine{\symRegRange}{q}
\symDefine{\symRegVal}{k}
\symDefine{\symRegValVariate}{K}
\symDefine{\symNumReg}{m}
\symDefine{\symDataItem}{D}
\symDefine{\symBitRepA}{a}
\symDefine{\symBitRepB}{b}
\symDefine{\symIndexI}{i}
\symDefine{\symIndexK}{k}
\symDefine{\symIndexL}{l}
\symDefine{\symIndexJ}{j}
\symDefine{\symCount}{c}
\symDefine{\symAlpha}{\alpha}
\symDefine{\symBeta}{\beta}
\symDefine{\symA}{a}
\symDefine{\symB}{b}
\symDefine{\symE}{e}
\symDefine{\symX}{x}
\symDefine{\symXEstimate}{\hat{\symX}}
\symDefine{\symXNormalized}{\symX'}
\symDefine{\symN}{n}
\symDefine{\symS}{s}
\symDefine{\symY}{y}
\symDefine{\symZ}{z}
\symDefine{\symInteger}{z}
\symDefine{\symIntegerVariate}{Z}
\symDefine{\symCardinality}{n}
\symDefine{\symCardinalityEstimate}{\hat{\symCardinality}}
\symDefine{\symCardinalityRawEstimate}{\symCardinalityEstimate_\text{raw}}
\symDefine{\symError}{\varepsilon}
\symDefine{\symStopDelta}{\delta}
\symDefine{\symStopEpsilon}{\varepsilon}
\symDefine{\symBigO}{\mathcal{O}}
\symDefine{\symExpectation}{\mathbb{E}}
\symDefine{\symProbability}{P}
\symDefine{\symProbabilityMass}{\rho}
\symDefine{\symRegProbability}{\gamma}
\symDefine{\symNu}{\nu}
\symDefine{\symKappa}{\kappa}
\symDefine{\symLikelihood}{\mathcal{L}}
\symDefine{\symPoissonRate}{\lambda}
\symDefine{\symPoissonRateEstimate}{\hat{\symPoissonRate}}
\symDefine{\symFunc}{f}
\symDefine{\symFuncPrime}{g}
\symDefine{\symHelper}{h}
\symDefine{\symHelperApprox}{\tilde{\symHelper}}
\symDefine{\symSetA}{A}
\symDefine{\symSetB}{B}
\symDefine{\symSetS}{S}
\symDefine{\symSetX}{X}
\symDefine{\symSetASuffix}{a}
\symDefine{\symSetBSuffix}{b}
\symDefine{\symSetXSuffix}{x}
\symDefine{\symCountMatrix}{\mathbf{c}}
\symDefine{\symPowerSeriesFunc}{\xi}
\symDefine{\symSmallCorrectionFunc}{\sigma}
\symDefine{\symLargeCorrectionFunc}{\tau}
\symDefine{\symEpsPowerSeriesFunc}{\varepsilon_\symPowerSeriesFunc}

@@

In the history of the HyperLogLog algorithm different averaging techniques have been proposed. First, there was the LogLog algorithm using the geometric mean and the SuperLogLog algorithm that enhanced the estimate by truncating the largest register values before applying the geometric mean \cite{Durand2003}. Finally, the harmonic mean was found to give even better estimates as it is inherently less sensitive to outliers. The result is the so-called raw estimator and given by
\begin{equation}
\label{equ:intermediate_estimate}
\symCardinalityRawEstimate
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexJ=1}^{\symNumReg}2^{-\symRegVal_\symIndexJ}}
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCount_\symRegVal 2^{-\symRegVal}}
\end{equation}
where $\symAlpha_\symNumReg$  is a correction factor that is fixed for a given number of registers $\symNumReg$ and which was derived to be \cite{Flajolet2007}
\begin{equation}
\symAlpha_\symNumReg := \left(
\symNumReg
\int_0^\infty
\left(
\log_2\left(
\frac{2+u}{1+u}
\right)
\right)^\symNumReg
du
\right)^{-1}.
\end{equation}
Numerical approximations of $\symAlpha_\symNumReg$ for various values of $\symNumReg$ have been listed in \cite{Flajolet2007}. These approximations are used in many HyperLogLog implementations. However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions $\symPrecision$. For HyperLogLog sketches that are used in practice with 256 or more registers ($\symPrecision\geq 8$), it is completely sufficient to use 
\begin{equation}
\label{equ:alpha_inf}
\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg = \frac{1}{2\log 2} \approx 0.7213475,
\end{equation}
as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:intermediate_estimate}, because the additional bias is negligible compared to the estimation error.

@@

As mentioned in \cref{sec:data_element_insertion}, a HyperLogLog sketch with parameters $(\symPrecision, \symRegRange)$ can be turned into a $(\symPrecision, 0)$-HyperLogLog sketch. Since $\symRegRange=0$ corresponds to linear counting and the transformed HyperLogLog sketch corresponds to a bitset with $\symCount_0$ zeros, the linear counting cardinality estimator \cite{Whang1990} can be applied here
\begin{equation}
\label{equ:small_range_estimate}
\symCardinalityEstimate_\text{small} = \symNumReg \log(\symNumReg/\symCount_0).
\end{equation}
The corresponding relative estimation error as depicted in \cref{fig:small_range_estimate} shows that this estimator is convenient for small cardinalities. It was proposed to use this small-range estimator as long as $\symCardinalityRawEstimate \leq \frac{5}{2}\symNumReg$ where the factor $\frac{5}{2}$ was empirically determined \cite{Flajolet2007}.

For large cardinalities in the order of $2^{\symPrecision+\symRegRange}$, for which a lot of registers are already in a saturated state meaning that they have reached the maximum possible value $\symRegRange + 1$, the raw estimator underestimates the cardinalities. For the 32-bit hash value case $(\symPrecision+\symRegRange=32)$ which was considered in \cite{Flajolet2007}, following correction formula was proposed to take these saturated registers into account
\begin{equation}
\label{equ:large_range_estimate}
\symCardinalityEstimate_\text{large}
=
-2^{32}\log\!\left(1-\symCardinalityRawEstimate/2^{32}\right).
\end{equation}
The original estimation algorithm as presented in \cite{Flajolet2007} including small and large range corrections is summarized by \cref{alg:estimate_original}. 
\begin{algorithm}
\caption{Original cardinality estimation algorithm for HyperLogLog sketches that use 32-bit hash values ($\symPrecision+\symRegRange = 32$) for insertion of data items \cite{Flajolet2007}.}
\label{alg:estimate_original}
\begin{algorithmic}
\Function {EstimateCardinality}{$\vec{\symRegVal}$}
\State $\symNumReg \gets \dim \vec{\symRegVal}$
\State $\symCardinalityRawEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexJ=1}^\symNumReg 2^{-\symRegVal_\symIndexJ}\right)^{-1}$
\Comment raw estimate \eqref{equ:intermediate_estimate}
\If{$\symCardinalityRawEstimate\leq \frac{5}{2}\symNumReg$}
\State $\symCount_0 = \left|\{\symIndexJ |\symRegVal_\symIndexJ=0\}\right|$
\If{$\symCount_0\neq0$}
\State \Return $\symNumReg \log(\symNumReg / \symCount_0)$ \Comment small range correction \eqref{equ:small_range_estimate}
\Else
\State \Return $\symCardinalityRawEstimate$
\EndIf
\ElsIf{$\symCardinalityRawEstimate\leq \frac{1}{30}2^{32}$}
\State \Return $\symCardinalityRawEstimate$
\Else
\State\Return $-2^{32}\log(1-\symCardinalityRawEstimate/2^{32})$\Comment large range correction \eqref{equ:large_range_estimate}
\EndIf
\EndFunction
\end{algorithmic}
\end{algorithm}
The relative estimation error for the original method is shown in \cref{fig:original_estimate}. Unfortunatley, as can be clearly seen, the ranges where the estimation error is small for $\symCardinalityRawEstimate$ and $\symCardinalityEstimate_\text{small}$ do not overlap. Therefore, the estimation error is much larger near the transition region. To reduce the estimation error for cardinalities close to this region, it was proposed to correct $\symCardinalityRawEstimate$ for bias. Empirically collected bias correction data is either stored as set of interpolation points \cite{Heule2013}, as lookup table \cite{Rhodes2015}, or as best-fitting polynomial \cite{Sanfilippo2014}. However, all these empirical approaches treat the symptom and not the cause.

@@

The large range correction is not satisfying either as it does not reduce the estimation error. Quite the contrary, it even increases the estimation error. However, instead of underestimating the cardinalities, they are now overestimated. Another indication for the incorrectness of the proposed large range correction \eqref{equ:large_range_estimate} is the fact that it is not even defined for all possible states. For instance, consider a $(\symPrecision,\symRegRange)$-HyperLogLog sketch with $\symPrecision+\symRegRange=32$ for which all registers are equal to the maximum possible value $\symRegRange+1$. The raw estimate would be $\symCardinalityRawEstimate = \symAlpha_\symNumReg 2^{33}$, which is greater than $2^{32}$ and outside of the domain of the large range correction.

A simple approach to avoid the need of any large range correction is to extend the operating range of the raw estimator to larger cardinalities. This can be accomplished by using hash values with more bits $(\symPrecision+\symRegRange>32)$. The drawback is that 5-bit registers are no longer sufficient to represent all possible values in the range $[0, \symRegRange+1]$, if $\symRegRange \geq 30$. For HyperLogLog sketches with 6-bit registers in combination with 64-bit hash values $(\symPrecision+\symRegRange=64)$ as proposed in \cite{Heule2013}, it is unrealistic in practice to encounter cardinalities of order $2^{64}$, which would require some correction.

@@

Consider following integer distribution
\begin{equation}
\label{equ:assumed_register_val_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate) = e^{-\frac{\symPoissonRate}{\symNumReg 2^{ \symRegVal}}}.
\end{equation}
This distribution has infinite support and differs from the register value distribution under the Poisson model \eqref{equ:register_value_distribution}, whose support is limited to the range $[0, \symRegRange+1]$. The expectation of $2^{-\symRegVal}$ is given by
\begin{multline}
\label{equ:expectation_power_two}
\symExpectation(2^{-\symRegVal})
=
\\
\sum_{\symRegVal = -\infty}^\infty
2^{-\symRegVal}
e^{-\frac{\symPoissonRate}{\symNumReg} 2^{-\symRegVal}}
\left(1-e^{-\frac{\symPoissonRate}{\symNumReg} 2^{-\symRegVal}}\right)
=
\frac{1}{2}
\sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal}
e^{-\frac{\symPoissonRate}{\symNumReg} 2^{\symRegVal}}
=
\frac{\symAlpha_\infty\,\symNumReg\,\symPowerSeriesFunc\!\left(\log_2\!\left(\symPoissonRate/\symNumReg\right)\right)}{\symPoissonRate},
\end{multline}
where the function 
\begin{equation}
\label{equ:power_series_function}
\symPowerSeriesFunc(\symX):= \log(2) \sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal+\symX}
e^{-2^{\symRegVal+\symX}}
\end{equation}
 is a smooth periodic function with period 1. Numerical evaluations indicate that this function can be bounded by $1 - \symEpsPowerSeriesFunc 
\leq \symPowerSeriesFunc(\symX) \leq 1 + \symEpsPowerSeriesFunc$ with $\symEpsPowerSeriesFunc:=9.885\cdot 10^{-6}$ (see \cref{fig:power_series_func}).

@@

Let $\symRegVal_1,\ldots,\symRegVal_\symNumReg$ be a sample taken from \eqref{equ:assumed_register_val_distribution}. For large sample sizes $\symNumReg\rightarrow \infty$ we have asymptotically
\begin{equation}
\symExpectation\!\left(
\frac{1}{2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}}
\right)
\underset{\symNumReg\rightarrow \infty}{=}
\frac{1}{\symExpectation
\!\left(2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}\right)}
=
\frac{1}{\symNumReg \symExpectation\!\left(2^{-\symRegVal}\right)}.
\end{equation}
Together with \eqref{equ:expectation_power_two} we obtain
\begin{equation}
\symPoissonRate
=
\symExpectation\!\left(
\frac{\symAlpha_\infty\,\symNumReg^2\,\symPowerSeriesFunc\!\left(\log_2\!\left( \symPoissonRate/\symNumReg\right)\right)}{2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}}
\right)
\quad
\text{for $\symNumReg\rightarrow \infty$}.
\end{equation}
Therefore, 
\begin{equation}
\label{equ:rawestimator_2}
\symPoissonRateEstimate 
= 
\frac{\symAlpha_\infty\,\symNumReg^2}{2^{-\symRegVal_1}+\ldots+2^{-\symRegVal_\symNumReg}}
\end{equation}
is an asymptotically almost unbiased estimator for the Poisson parameter. Its asymptotic relative bias is bounded by $\symEpsPowerSeriesFunc$. This estimator corresponds to the raw estimator \eqref{equ:intermediate_estimate}, if the Poisson parameter estimate is used as cardinality estimate (see \cref{sec:poisson_approximation}).

@@

\subsection{Limitations of the raw estimator}
As we have already seen in \cref{fig:raw_estimate}, the raw estimator does not work very well for small or large cardinalities. The reason is that the distribution of register values \eqref{equ:register_value_distribution} is different from \eqref{equ:assumed_register_val_distribution} for which the raw estimator was derived. Since a random variable $\symRegValVariate'$ that obeys \eqref{equ:assumed_register_val_distribution} can be transformed into random variable $\symRegValVariate$ that follows \eqref{equ:register_value_distribution} using the transformation $\symRegValVariate = \min\left(\max\left(\symRegValVariate', 0\right), \symRegRange+1\right)$, the register values of a HyperLogLog sketch can be seen as the result after applying this transformation to a sample that is distributed according to \eqref{equ:assumed_register_val_distribution}. The corresponding multiplicities have  following relationships
\begin{equation}
\label{equ:multiplicity_transformation}
\begin{aligned}
\symCount_0 &= \textstyle\sum_{\symRegVal = -\infty}^{0} \symCount'_\symRegVal, \\
\symCount_\symRegVal &=  \symCount'_\symRegVal, \quad 1\leq\symRegVal\leq\symRegRange,\\
\symCount_{\symRegRange+1} &= \textstyle\sum_{\symRegVal = \symRegRange + 1}^{\infty} \symCount'_\symRegVal.
\end{aligned}
\end{equation}
Obviously, as long as $\symCount_0\ll\symNumReg$ and  $\symCount_{\symRegRange+1}\ll\symNumReg$ the set of register values is similar to the originating sample. In case $\symCount_0=0$ and $\symCount_{\symRegRange+1}=0$, which implies $\symCount'_\symRegVal=0$ for $\symRegVal\leq0$ and $\symRegVal\geq\symRegRange+1$, we can even be sure that the register values are identical to the originating sample. This explains why the raw estimator only works well in the intermediate cardinality range for which $\symCount_0$ and $\symCount_{\symRegRange+1}$ are negligible.

\subsection{Corrections to the raw estimator}
Given a HyperLogLog sketch with multiplicity vector $\vec{\symCount}=\left(\symCount_0,\ldots,\symCount_{\symRegRange+1}\right)$, we try to find estimates $\hat{\symCount}'_\symRegVal$ for $\symCount'_\symRegVal$ for all $\symRegVal\in\mathbb{Z}$ and replace them in the estimation formula \eqref{equ:raw_estimate_with_multiplicities}. For $\symRegVal\in[1,\symRegRange]$ where $\symCount_\symRegVal = \symCount'_\symRegVal$ we can use the trivial estimators 
\begin{equation}
\hat{\symCount}'_\symRegVal := \symCount_\symRegVal,\quad 1\leq\symRegVal\leq\symRegRange.
\end{equation}
To get estimators for all other $\symRegVal$, we consider their expectations
\begin{equation}
\symExpectation(\symCount'_\symRegVal)
=
\symNumReg e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal}}}\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal}}}\right).
\end{equation}
From \eqref{equ:register_value_distribution} we know that $\symExpectation\!\left(\symCount_0/\symNumReg\right)=
e^{-\frac{\symPoissonRate}{\symNumReg}}$ and $\symExpectation\!\left(1-\symCount_{\symRegRange+1}/\symNumReg\right)=
e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}$, and therefore, we can also write
\begin{equation}
\symExpectation(\symCount'_\symRegVal)
=
\symNumReg
\left(\symExpectation\!\left(\symCount_0/\symNumReg\right)\right)^{2^{-\symRegVal}}
\left(1-\left(\symExpectation\!\left(\symCount_0/\symNumReg\right)\right)^{2^{-\symRegVal}}\right)
\end{equation}
and
\begin{equation}
\symExpectation(\symCount'_\symRegVal)
=
\symNumReg
\left(\symExpectation\!\left(1-\symCount_{\symRegRange+1}/\symNumReg\right)\right)^{2^{\symRegRange-\symRegVal}}
\left(1-\left(\symExpectation\!\left(1-\symCount_{\symRegRange+1}/\symNumReg\right)\right)^{2^{\symRegRange-\symRegVal}}\right),
\end{equation}
which motivates us to use
\begin{equation}
\hat{\symCount}'_\symRegVal
:=
\symNumReg\cdot
\left(\symCount_0/\symNumReg\right)^{2^{-\symRegVal}}\cdot
\left(1-\left(\symCount_0/\symNumReg\right)^{2^{-\symRegVal}}\right)
\end{equation}
as estimator for $\symRegVal \leq 0$ and
\begin{equation}
\hat{\symCount}'_\symRegVal
:=
\symNumReg\cdot
\left(1-\symCount_{\symRegRange+1}/\symNumReg\right)^{2^{\symRegRange-\symRegVal}}
\cdot
\left(1-\left(1-\symCount_{\symRegRange+1}/\symNumReg\right)^{2^{\symRegRange-\symRegVal}}\right)
\end{equation}
as estimator for $\symRegVal \geq \symRegRange+1$, respectively.
